On the Utility of Azimuth and Decentralized Identity

### 2024-08-29 mss.tex
Post after conversion and review.

\usepackage{ustj}

@@

\newcommand{\authorname}{Stuart Christoph}
\newcommand{\authorpatp}{\patp{lagrev-nocfep}}
\newcommand{\affiliation}{\patp{tocwex} Syndicate}

@@

%%MANUSCRIPT
\title{Ruminations on the Utility of Azimuth and Decentralized Identity}
\author{\authorname~\authorpatp \\ \affiliation}
\date{}

@@

\begin{abstract}
\end{abstract}

@@

Blockchain developer Christopher Allen coined the term `self-sovereign identity' (\textsc{ssi}) to concretize a notion of digitial decentralized identity built for individual control and self-determination \citep{Allen2016}. He noted that the evolution of online identity has passed through eras of centralized administrative control, oligarchic federated systems, and unfortunately, institutionally captured ``user-centric'' designs. The next era, designated the era of \textsc{ssi}, thankfully aims to return us to a more intuitive understanding of digital identity. Allen defined ten key principles for \textsc{ssi}: \footnote{The appendix contains an excerpt defining the principles more completely}.
% Allen2016 https://www.lifewithalacrity.com/article/the-path-to-self-soverereign-identity/

\begin{tabular}{ll}
  \begin{itemize}
    \item  Existence
    \item  Control
    \item  Access
    \item  Transparency
    \item  Persistence
  \end{itemize}
  &
  \begin{itemize}
    \item  Portability
    \item  Interoperability
    \item  Consent
    \item  Minimalization
    \item  Protection
  \end{itemize}
\end{tabular}

@@

Separate from Allen's principles, W3C has presented a standard for Decentralized Identifiers (\textsc{did}s). According to ``Decentralized Identifiers (\textsc{did}s) v1.0", in part \citep{W3C2022}:
% (ref https://www.w3.org/TR/did-core/)

\begin{quote}
[Decentralized identifiers](https://www.w3.org/TR/did-core/#dfn-decentralized-identifiers) (\textsc{did}s) are a new type of identifier that enables verifiable, decentralized digital identity. A [\textsc{did}](https://www.w3.org/TR/did-core/#dfn-decentralized-identifiers) refers to any subject (e.g., a person, organization, thing, data model, abstract entity, etc.) as determined by the controller of the [\textsc{did}](https://www.w3.org/TR/did-core/#dfn-decentralized-identifiers).
\ldots{}
While other parties might be used to help enable the discovery of information related to a [\textsc{did}](https://www.w3.org/TR/did-core/#dfn-decentralized-identifiers), the design enables the controller of a [\textsc{did}](https://www.w3.org/TR/did-core/#dfn-decentralized-identifiers) to prove control over it without requiring permi\textsc{ssi}on from any other party. [\textsc{did}s](https://www.w3.org/TR/did-core/#dfn-decentralized-identifiers) are [URIs](https://www.w3.org/TR/did-core/#dfn-uri) that associate a [\textsc{did} subject](https://www.w3.org/TR/did-core/#dfn-did-subjects) with a [\textsc{did} document](https://www.w3.org/TR/did-core/#dfn-did-documents) allowing trustable interactions associated with that subject.
\end{quote}

While the Urbit project may not traditionally rely on `Earth standards,' the \textsc{did} recommendation also defines ten design goals that are informative for understanding decentralized identity. (FOOTNOTE: See section 6: Appendix for full excerpt defining the design goals):

\begin{tabular}{ll}
  \begin{itemize}
    \item  Decentralization
    \item  Control
    \item  Privacy
    \item  Security
    \item  Proof-based
  \end{itemize}
  &
  \begin{itemize}
    \item  Discoverability
    \item  Interoperability
    \item  Portability
    \item  Simplicity
    \item  Extensibility
    \end{itemize}
\end{tabular}

@@

We understand the need to place the \textsc{pki} on a distributed ledger with global consensus and recognize the hierarchical model as a sane default. But what do we know about the current address space and its technical implementation? Since 2017, Azimuth has been on hosted on the Ethereum blockchain, and in 2022 the contracts were updated to support a ``naive'' Layer 2 rollup. Returning to the principles and design goals of \textsc{ssi} and \textsc{did}s, this on-chain presence provides benefits such as existence, control, transparency, persistence, decentralization, and discoverability. It also supports many of the other stated goals, if not satisfying them purely through an on-chain presence. Many of these principles were either explicitly or implicitly considered with both the initial Azimuth contracts (ref blog post from midden fabler), and the move to supporting the Layer 2 rollup (ref blog post about solving the gas crisis). However, the current state of Azimuth falls short in terms of providing interoperability and extensibility.

@@

The deployment of Azimuth on Ethereum was one of the first uses of the EIP-721 standard (ref https://eips.ethereum.org/EIPS/eip-721). It is implemented across two core contracts: Azimuth, which holds the registry of identity ownership, and Ecliptic, which defines the business logic for updating the state of the registry. To this day, these contracts serve as one of the most technically `useful' ERC-721s on the Ethereum network, powering the peer-to-peer interactions of thousands of nodes (also known as urbit ships) in an encrypted off-chain network. As an ERC-721, Urbit ID experienced a significant increase in valuation during the NFT boom of 2021, with the market cap of address space breaking \$2B at its peak.

@@

The transaction costs on Ethereum were so great that, at times, it cost over \$200 worth of ETH just to get on the network. For a budding distributed system project, such costs can be unbearable. Thus, the `solution to the gas crisis' (ref https://urbit.org/blog/rollups) was introduced: a custom `naive' rollup. The key feature of the naive rollup is that, instead of paying Ethereum miners (now validators) to verify the state transitions of Azimuth, each Urbit node can validate those transitions themselves. This again brings us back to the idea of sane defaults in a decentralized or distributed network.

That is, in running your urbit, the canonical ``Azimuth" is nothing more than a sane default. This is exemplified by the fact that you could run a `fakezod' network locally and act as any Urbit ID you chose, or even deploy an alternative version of Azimuth onto mainnet Ethereum. However, because everyone else is likely operating on the default version of `azimuth.hoon`, which assumes `azimuth.eth` as The One True \textsc{pki}, any of your potential peers will discard your packets as invalid. The `Urbit network' will ignore you. Whether or not the Urbit network pays the Ethereum network for state transition computation, it remains true that Urbit's off-chain interactions are voluntary and governed by sane defaults.

@@

As noted in `~wicdev-wisryt`'s piece, blockchains allow us to distinguish between two cryptographically valid messages, specifically which was signed first, without requiring a centralized party. Both the Layer 1 and Layer 2 implementations rely on this affordance. Effectively, they both need the chain to give them a timestamp and a signed message to discern key ownership. Whether the computation of key rotation is done on-chain or client side, anyone within the Urbit network that is continuing to use the sane defaults the system provides will ultimately get the correct result for validating a packet's sender. It is the core affordance of timestamped messages for which Urbit must pay *some* blockchain to have a viable decentralized \textsc{pki}.

However, Urbit-side key exchange and peer discovery are not the only benefits of having the \textsc{pki} on Ethereum. The cost of on-chain computation for global consensus is undoubtedly a hurdle to low-cost user adoption of Urbit. However, it would be a failure of imagination to claim there is no contingent loss of benefits. Specifically, Ethereum is a growing network that is building on top of the same ERC-721 standard on which Urbit ID is defined. Improvements to the capabilities of the Ethereum network can enhance the capabilities of the Urbit network—*if Urbit maintains, or increases, its legibility to the Ethereum Network*.  The current naive rollup does not achieve this. An L2 Urbit ID, and it's ownership in particular, is functionally illegible to the outside world. From the perspective of Ethereum, global consistency is compromised, and key rotation is impo\textsc{ssi}ble to validate.

While we could explore the technical details of these two systems, instead let's return to examine the \textsc{ssi} principles and \textsc{did} design goals of interoperability and extensibility. In a sense, these just mean that you want your identity to work in as many places as po\textsc{ssi}ble. To be as legible to as many systems or stakeholders as is po\textsc{ssi}ble.

A playful example of this is whether you would trade your state-issued driver's license for a high school ID card (ref https://sarlev-sarsen.rooftopdao.io/blog/why-l1masterrace-is-more-than-a-meme). In spite of your driver's license being more costly to replace or update, your response was almost certainly, ``No way." You know intuitively that your driver's license is a more credible and legible identifier than a high school ID. Now think if you could achieve that level of credibility and legibility in a self-sovereign manner.

But this tradeoff isn't just a fun thought experiment; a dependency on centralized identity issuance authorities has real world impacts:

> ``In the last year, self-sovereign identity has also entered the sphere of _international policy_ [17](https://blogs.worldbank.org/en/digital-development/world-citizen-transforming-statelessness-global-citizenship). This has largely been driven by the refugee crisis that has beset Europe, which has resulted in many people lacking a recognized identity due to their flight from the state that issued their credentials. However, it’s a long-standing international problem, as foreign workers have often been abused by the countries they work in due to the lack of state-issued credentials." (ref https://www.lifewithalacrity.com/article/the-path-to-self-soverereign-identity/)

Now, we can ask, `If our Urbit IDs are going to be the foundation for a self-sovereign digital identity, do we want them to be more legible or less legible?'

@@

Improving legibility, or in other words, protecting the interoperability and extensibility of Urbit ID, does not imply loyalty to a particular chain. It simply claims that increasing legibility to the largest set of networks should be included in the objective function of Urbit ID.

@@

\How do we continue to improve Azimuth? While in 2017 the options for the \textsc{pki} were limited, the technology landscape has drastically evolved in the intervening years.

\subsection{Self-Host the \textsc{pki} on \lstinline[style=inlinecode]{%chain}}
\
The most `philosophically straightforward' answer would be to self-host the \textsc{pki}. While this requires further technical development, it is conceptually tractable and the \lstinline[style=inlinecode]{%chain} project by `~tiller-tolbus` and `~midsum-salrux` has shown initial successes on an urbit-native blockchain. This option could make use of identity-backed consensus models, and would be the most direct pathway to the Urbit Network not having to pay a `tax' to another network for its consensus mechanism. In a sense, this is the most `independent' or `self-contained' option. However, it would likely result in decreased capital access and practical legibility to external networks.

@@

Moving the \textsc{pki} in this way would also decrease interoperability and extensibility by making it harder for the Ethereum network to access and validate data about ownership of Urbit IDs. Technically speaking, this constraint could be reduced by self-hosted RPC endpoints of \lstinline[style=inlinecode]{%chain} or Azimuth state that could be accessed over HTTP. However, the more prominent issue would be the coordination constraint it would generate. If the \textsc{pki} moves, what motivation do members of the Ethereum Network have to *bother* with Urbit RPC endpoints instead of just staying within the Ethereum tooling ecosystem?

In a world where Urbit is more powerful and offers something compelling to Ethereum, you could see the draw. However, the ~70k per year in gas fees that Urbit currently pays to maintain the \textsc{pki} on Ethereum seems worth the legibility it affords.

\subsection{Zenith:  UrbitChain on Cosmos \texsc{sdk}}

Of course, moving the \textsc{pki} to a self-hosted, urbit-native solution isn't the only option if the Urbit Network wants to reduce its dependency on outside networks. For example, there are ongoing discu\textsc{ssi}ons about an `Urbitchain,' governed by the galaxy-level address space and implemented as its own chain using the Cosmos \textsc{sdk}.

In terms of legibility, this pathway essentially splits the difference. The Cosmos ecosystem is smaller than Ethereum as a whole—coming in at #2 on metrics for Total Value Locked (TVL) (ref https://thedefiant.io/news/markets/cosmos-tvl-surge-ethereum) —but could give some of the benefits of a self-hosted system. For example, it would reduce the outflow of capital from the Urbit network, while still achieving some interoperability with the external world thanks to the affordances granted by Cosmos. Peg zones and the Inter-blockchain Communication (IBC) protocol would mean access to a network of blockchain networks. These networks are likely to have more operational friction for capital flows than something like remaining on L1 Ethereum, but may be a valid tradeoff.

\subsection{Split the \textsc{pki} across multiple chains}

As we consider the tradeoffs of moving the \textsc{pki}, it is worth revisiting our `sane defaults' branch in the idea maze. If `azimuth.eth` and with it `azimuth.hoon`, are just sane defaults, why not also support some sane *alternatives*? One of the final Urbit precepts (ref Urbit precepts) from  `~wicdev-wisryt` is `Communities are autonomous.'

@@

Some may choose to remain on mainnet Ethereum to take advantage of features like ERC-6551 Token Bound Accounts(ref https://eips.ethereum.org/EIPS/eip-6551) (FOOTNOTE: EIP-6551 implements an interface and registry for smart contract accounts deterministically controlled by your ERC-721 Urbit ID). Others may want to inscribe their ownership as an Ordinal on Bitcoin. Alternatively, Solana offers some benefits for the more crypto-degenerate who are looking for speed of trading and market volatility to help make their wealth.

Considering this path, it is worthwhile to elaborate on how the current L2 solution works. The original version of the Ethereum contracts held ownership data in the Azimuth contracts and business logic in the Ecliptic contracts. In 2021, these contracts were updated to support essentially two new affordances:

\begin{itemize}
  \item  Azimuth points sent to the ethereum deposit address (`0x1111...`) are assessed by the default Urbit client as being owned on layer 2.
  \item  Ecliptic includes a function (TK, check where this function actually lives? Open questions out to `~norsyr-torryn`) for posting *time-stamped* signature data to the Ethereum chain concerning the change of ownership of a point owned on layer 2, and the default urbit client will review and validate the state change before applying the change to the urbit's own `azimuth.hoon`.
\end{itemize}

This mechanism means that these points are still cryptographically controlled for purposes of anyone running the default Urbit OS or relevant code (namely, `eth-watcher.hoon`, `naive.hoon`, `azimuth-tracker.hoon` and `azimuth.hoon`), and it is easy to imagine how this same mechanism could be used to support other chains as hosts for portions of the \textsc{pki}:
1. Define a holding address for the ERC-721 on mainnet ethereum for other host chains (i.e. Sending the L1 NFT to `0x2222...` indicates sending the ownership to bitcoin)
2. Implement a watcher for the subject chain that looks for posted signature data
3. Update and/or add to the relevant Urbit OS code for client side validation of data posted to the new chain.
This approach would increase legibility to other chains, giving communities more options for what their members can achieve with their Urbit ID.

However, as the \textsc{pki} fractures, maintaining global consistency becomes increasingly complex. In the above model, the primary solution for maintaining this consistency becomes that official `global consensus' is on mainnet Ethereum, while any other alternative is subject only to the client side validation. Depending on the affordances of the subject chain, the level of effort put into building the subsystem, and any cost concerns for network fees on the specific subject chain, the result is likely that *moving to any other chain is a one-way trip*. Just as a move to the current naive rollup is a one way trip.

@@

\subsection{Move the \textsc{pki} to Nockchain}

While Ethereum has made significant progress in interchain bridges and L2 rollups since the implementation of the Naive rollup, making two-way bridges increasingly po\textsc{ssi}ble, they still introduce a host of security and consensus questions. They also are likely to exacerbate the capital outflow concerns of the current L1 Ethereum implementation.

@@

What is the best way to enable this portability (another shared principle/design goal for \textsc{ssi}s/\textsc{did}s) while maintaining global consistency? If you read the section title, and know what Nockchain is, then you will know what I am about to say:

ZK Proofs.

Using Zorp's EDEN and associated Nock prover (ref https://eprint.iacr.org/2023/1021), any Urbit could prove valid computation of any `client-side' Azimuth state change and use that proof as the data to post to any given blockchain. It is likely that the `easiest' place to do that would be Nockchain. However, external legibility might incentivize these proofs to be posted elsewhere. An Azimuth contract on a subject chain could use the proof of valid nock computation to a\textsc{ssi}gn ownership of an Urbit ID crypto-asset native to a given address. This would include whatever affordances native asset ownership on that chain may bestow to an Urbit operator.

In the long run, this pathway would likely lead more Urbit operators to move their Urbit ID's primary `citizenship' to Nockchain. This would occur without unnecessarily losing legibility to other networks and with the po\textsc{ssi}bility of absorbing external capital into the Urbit network along the way.

\section{Other Considerations for Azimuth as a decentralized identity system}

While Azimuth's on-chain nature is that of a public key infrastructure, a \textsc{pki} on it's own doesn't make for a complete identity system. In fact, it is the pairing of Urbit ID with Urbit OS that presents the opportunity to serve as a full-fledged decentralized identity system. Whether we look at implementing W3C's standardized \textsc{did}s, or adhering to the principles of \textsc{ssi}s, not everything about Azimuth is, or should be, on-chain.

@@

Globally available and transparent chain state greatly enhances legibility. Openly available sponsorship hierarchies inform unconnected peers about how to reach you, and global claims can be used for various purposes. However, not everything needs global consensus (ref [[How much consensus? How smart of contracts?]]). In fact, many networks may want to be illegible to the outside. Private access control lists, go\textsc{ssi}p networks, and packet routing mechanisms enable use cases that would not benefit from the cost or transparency of on-chain or globally consistent legibility. Urbit's off-chain and encrypted peer-to-peer messaging affordances open up the landscape of `illegible subnets' while still benefiting from associations with specific pseudonymous urbit identities.

The value of illegible subnets also builds on the insight that not everything can, or should, be distilled down to its lowest common denominator, tokenized, and put on the blockchain. The more complex the interaction, the more costly the distillation process—and *the more likely there is no consensus to be found around its `valuation'*. Instead of demanding timestamped global consensus in all our peer interactions, we must support emergent, implied-consensus networks. These networks are governed not by immutable chain state but by willing association and participation by sovereign peers. Systems where the network forms and dissolves on the fly, based on the ongoing discovery and refinement of the values of its participants.

It is worth noting that illegible subnets may contend with the \textsc{ssi} principle of consent, in that they enable opaque data-sharing networks and other similar `darknet' use cases.

@@

The principles of \textsc{ssi} are as follows. An excerpt from the section titled ``Ten Principles of Self-Sovereign Identity," published April 26, 2016:

>>>>

\usepackage{ustj}
\usepackage{booktabs}% http://ctan.org/pkg/booktabs
\newcommand{\tabitem}{~~\llap{\textbullet}~~}

@@

\newcommand{\authorname}{Stuart Christoph}
\newcommand{\authorpatp}{\patp{sarlev-sarsen}}
\newcommand{\affiliation}{\texttt{\textasciitilde tocwex.syndicate}}

@@

%%MANUSCRIPT
\title{On the Utility of Azimuth and Decentralized Identity}
\author{\authorname~\authorpatp \\ \affiliation}
\date{}

@@

\begin{abstract}
  This article considers the concept of identity within the context of Urbit and its identity layer, Azimuth.  Comparison is made to \textsc{ssi} and \textsc{did} design principles.  We consider how these systems implement decentralized identity, examining their technical foundations, practical applications, and the broader implications for digital self-sovereignty.  Understanding these elements sheds light on the future of identity in a decentralized world.
\end{abstract}

@@

Blockchain developer Christopher Allen coined the term `self-sovereign identity' (\textsc{ssi}) to concretize a notion of digitial decentralized identity built for individual control and self-determination \citep{Allen2016}. He noted that the evolution of online identity has passed through eras of centralized administrative control, oligarchic federated systems, and unfortunately, institutionally captured ``user-centric'' designs. The next era, designated the era of \textsc{ssi}, thankfully aims to return us to a more intuitive understanding of digital identity. \citeauthor{Allen2016} defined ten key principles for \textsc{ssi}:\footnote{The appendix contains an excerpt defining the principles more completely}.

\begin{tabular}{ll}
  \tabitem  Existence
  &
  \tabitem  Portability \\
  \tabitem  Control
  &
  \tabitem  Interoperability \\
  \tabitem  Access
  &
  \tabitem  Consent \\
  \tabitem  Transparency
  &
  \tabitem  Minimalization \\
  \tabitem  Persistence
  &
  \tabitem  Protection \\
\end{tabular}

@@

Separate from Allen's principles, W3C has presented a standard for Decentralized Identifiers (\textsc{did}s). According to ``Decentralized Identifiers (\textsc{did}s) v1.0", in part \citep{W3C2022}:

\begin{quote}
Decentralized identifiers (\textsc{did}s) are a new type of identifier that enables verifiable, decentralized digital identity. A \textsc{did} refers to any subject (e.g., a person, organization, thing, data model, abstract entity, etc.) as determined by the controller of the \textsc{did}.  \ldots{}

While other parties might be used to help enable the discovery of information related to a \textsc{did}, the design enables the controller of a \textsc{did} to prove control over it without requiring permi\textsc{ssi}on from any other party. \textsc{did}s are \textsc{uri}s that associate a \textsc{did} subject with a \textsc{did} document allowing trustable interactions associated with that subject.
\end{quote}

While the Urbit project may not traditionally rely on `Earth standards,' the \textsc{did} recommendation also defines ten design goals that are informative for understanding decentralized identity.\footnote{See the appendix for a more complete excerpt explicating the design goals}

\begin{tabular}{ll}
  \tabitem  Decentralization
  &
  \tabitem  Discoverability \\
  \tabitem  Control
  &
  \tabitem  Interoperability \\
  \tabitem  Privacy
  &
  \tabitem  Portability \\
  \tabitem  Security
  &
  \tabitem  Simplicity \\
  \tabitem  Proof-based
  &
  \tabitem  Extensibility \\
\end{tabular}

@@

The need to place the \textsc{pki} on a distributed ledger with global consensus and recognize the hierarchical model as a sane default is obvious. But what about the current address space and its technical implementation? Since early 2019, Azimuth has been on hosted on the Ethereum blockchain, and in 2022 the contracts were updated to support a ``naive'' Layer 2 rollup. Returning to the principles and design goals of \textsc{ssi} and \textsc{did}s, this on-chain presence provides benefits such as existence, control, transparency, persistence, decentralization, and discoverability. It also supports many of the other stated goals, if not satisfying them purely through an on-chain presence. Many of these principles were either explicitly or implicitly considered with both the initial Azimuth contracts \citep{WolfePaul2019}, and the move to supporting the Layer 2 rollup \citep{Paprocki2021}. However, the current state of Azimuth falls short in terms of providing interoperability and extensibility.

@@

Azimuth's deployment on Ethereum was one of the first uses of the \textsc{erc}-721 standard \citep{ERC721}. It is implemented across two core contracts: Azimuth, which holds the registry of identity ownership, and Ecliptic, which defines the business logic for updating the state of the registry. To this day, these contracts serve as one of the most technically `useful' ERC-721s on the Ethereum network, powering the peer-to-peer interactions of thousands of nodes (also known as urbit ships) in an encrypted off-chain network. As an ERC-721, Urbit ID experienced a significant increase in valuation during the NFT boom of 2021, with the market cap of address space breaking \$2B at its peak.

@@

The transaction costs on Ethereum were so great that, at times, it cost over \$200 worth of \textsc{eth} just to get on the network. For a budding distributed system project, such costs can be unbearable. Thus, the solution to the `gas crisis' was introduced \citep{Paprocki2021}: a custom `naive' rollup. The key feature of the naive rollup is that, instead of paying Ethereum miners (now validators) to verify the state transitions of Azimuth, each Urbit node can validate those transitions themselves. This again brings us back to the idea of sane defaults in a decentralized or distributed network.

That is, in running one's Urbit instance, the canonical ``Azimuth" is nothing more than a sane default. This is trivially exemplified by the fact that one could run a fake ship network locally and act as any Urbit ID one chose, or even deploy an alternative version of Azimuth onto mainnet Ethereum. However, because everyone else is likely operating on the default version of \lstinline[style=inlinecode]{/app/azimuth}, which assumes \lstinline[style=inlinecode]{azimuth.eth} as The One True \textsc{pki}, any of one's potential peers will discard one's packets as invalid. The Urbit network will ignore an unsignable interloper. Whether or not the Urbit network pays the Ethereum network for state transition computation, it remains true that Urbit's off-chain interactions are voluntary and governed by sane defaults.

@@

As noted in \patp{wicdev-wisryt}'s piece, blockchains allow users to distinguish two cryptographically valid messages, specifically which was signed first, without requiring a centralized party. Both Layer 1 and Layer 2 implementations rely on this affordance. Effectively, they both need the chain to give them a timestamp and a signed message to discern key ownership. Whether the computation of key rotation is done on-chain or client side, anyone within the Urbit network that is continuing to use the sane defaults the system provides will ultimately get the correct result for validating a packet's sender. It is the core affordance of timestamped messages for which Urbit must pay \emph{some} blockchain to have a viable decentralized \textsc{pki}.

However, Urbit-side key exchange and peer discovery are not the only benefits of having the \textsc{pki} on Ethereum. The cost of on-chain computation for global consensus is undoubtedly a hurdle to low-cost user adoption of Urbit. However, it would be a failure of imagination to claim there is no contingent loss of benefits. Specifically, Ethereum is a growing network that is building on top of the same \textsc{erc}-721 standard on which Urbit ID is defined. Improvements to the capabilities of the Ethereum network can enhance the capabilities of the Urbit network—\emph{if Urbit maintains, or increases, its legibility to the Ethereum Network}.  The current naive rollup does not achieve this. An L2 Urbit ID, and it's ownership in particular, is functionally illegible to the outside world. From the perspective of Ethereum, global consistency is compromised, and key rotation is impossible to validate.

While we could explore the technical details of these two systems, instead let's return to examine the \textsc{ssi} principles and \textsc{did} design goals of interoperability and extensibility. In a sense, these just mean that you want your identity to work in as many places as possible. To be as legible to as many systems or stakeholders as is po\textsc{ssi}ble.

A playful example of this is whether one would trade one's state-issued driver's license for a high school ID card \citep{Sarlev2024}.  In spite of one's driver's license being more costly to replace or update, one's response is almost certainly, ``No way." A licensee knows intuitively that a driver's license is a more credible and legible identifier than a high school ID, in large part due to the authority of the centralized issuing authority. Imagine achieving such a level of credibility and legibility in a self-sovereign manner.  This tradeoff is not merely a thought experiment, as dependency on centralized identity issuance authorities has real world impacts:

\begin{quote}
In the last year, self-sovereign identity has also entered the sphere of \emph{international policy}. This has largely been driven by the refugee crisis that has beset Europe, which has resulted in many people lacking a recognized identity due to their flight from the state that issued their credentials. However, it’s a long-standing international problem, as foreign workers have often been abused by the countries they work in due to the lack of state-issued credentials.  \citep{Allen2016}
\end{quote}

One may reasonably ask at this junction whether, if Urbit IDs will form the foundation for self-sovereign digital identity, they should be more legible or less legible?

@@

Improving legibility, or in other words, protecting the interoperability and extensibility of Urbit ID, does not imply loyalty to any particular chain. It simply claims that increasing legibility to the largest set of networks should be included in the objective function of Urbit ID.

@@

How does Urbit continue to improve the utility of Azimuth? While in 2019 the options for hosting the \textsc{pki} were still limited, the technology landscape has drastically evolved in the intervening years.  Several Urbit-related blockchain projects have been proposed and explored recently which bear on how the \textsc{pki} could come to be hosted.

\subsection[Self-Host the \textsc{pki} on \texttt{\%chain}]{Self-Host the PKI on \texttt{\%chain}}

The most `philosophically straightforward' answer would be for Urbit to self-host its \textsc{pki}. While this requires further technical development, it is conceptually tractable and the \lstinline[style=inlinecode]{%chain} project, primarily executed by \patp{tiller-tolbus} and \patp{midsum-salrux} as a skunkworks effort, has shown initial successes on an Urbit-native blockchain. This option could make use of identity-backed consensus models, and would be the most direct pathway to the Urbit network not having to pay a `tax' to another network for its consensus mechanism. In a sense, this is the most `independent' or `self-contained' option. However, it would likely result in decreased capital access and practical legibility to external networks.

@@

Moving the \textsc{pki} in this way would also decrease interoperability and extensibility by making it harder for the Ethereum network to access and validate data about ownership of Urbit IDs. Technically speaking, this constraint could be reduced by self-hosted \textsc{rpc} endpoints of \lstinline[style=inlinecode]{%chain} or Azimuth state that could be accessed over \textsc{http}. However, the more prominent issue would be the coordination constraint it would generate. If the \textsc{pki} moves, what motivation do members of the Ethereum Network have to bother with Urbit \textsc{rpc} endpoints instead of just staying within the Ethereum tooling ecosystem?

In a world in which Urbit is more powerful and offers a compelling, market-legible alternative to the Ethereum virtual machine, one can see the draw of a bootstrapped \lstinline[style=inlinecode]{%chain}. However, the \textasciitilde 70K \textsc{usd} per year in gas fees that Urbit users collectively pay to maintain the \textsc{pki} on Ethereum seems worth the legibility it affords.

\subsection[Zenith:  UrbitChain on Cosmos \textsc{sdk}]{Zenith:  UrbitChain on Cosmos SDK}

Moving the \textsc{pki} to a self-hosted, urbit-native solution is not the only option should the Urbit network want to reduce its dependency on outside networks. One standing proposal, Zenith, establishes an off-Urbit blockchain governed by the galaxy-level address space and implemented as its own L1 chain using the Cosmos \textsc{sdk}.

In terms of legibility, this pathway essentially splits the difference. The Cosmos ecosystem is smaller than Ethereum as a whole, but comes in second place on metrics for current Total Value Locked (\textsc{tvl}) \citep{CosmosTVL}.  The Cosmos ecosystem and \textsc{sdk} could give some of the benefits of a self-hosted system. For example, it would reduce the outflow of capital from the Urbit network, while still achieving some interoperability with the external world thanks to the affordances granted by Cosmos. Peg zones and the Inter-Blockchain Communication (\textsc{ibc}) protocol would enable ready access to a network of blockchain networks. These networks are likely to have more operational friction for capital flows than remaining on L1 Ethereum, but may be a valid tradeoff.

\subsection[Bridge the \textsc{pki} across multiple chains]{Bridge the PKI across multiple chains}

As we consider the tradeoffs of moving the \textsc{pki}, it is worth revisiting our `sane defaults' branch in the idea maze. If \lstinline[style=inlinecode]{azimuth.eth} and with it \lstinline[style=inlinecode]{/app/azimuth.hoon}\footnote{The Jael secrets vane of Urbit OS merely subscribes for the \textsc{pki} state, rather than storing it in Arvo itself.} are just sane defaults, why not also support some sane \emph{alternatives}? One of the final Urbit precepts reminds us that ``communities are autonomous'' \citep{UrbitPrecepts}.

@@

Some may choose to remain on mainnet Ethereum to take advantage of features like \textsc{erc}-6551 Token Bound Account \citep{EIP6551}.\footnote{\textsc{erc}-6551 implements an interface and registry for smart contract accounts deterministically controlled by one's \textsc{erc}-721 Urbit ID.} Others may want to inscribe their ownership as an Ordinal on Bitcoin. Alternatively, Solana offers some benefits for the more crypto-degenerate who are looking for speed of trading and market volatility to help make their wealth.

Considering this path, it is worthwhile to elaborate on how the current L2 solution works. The original version of the Ethereum contracts held ownership data in the Azimuth contracts and business logic in the Ecliptic contracts. In 2021, these contracts were updated to support two new affordances:

\begin{itemize}
  \item  Azimuth points sent to the Ethereum deposit address (\lstinline[style=inlinecode]{0x1111...}) are assessed by the default Urbit client as being owned on layer 2.
  \item  Ecliptic posts \emph{time-stamped} signature data to the Ethereum chain concerning the change of ownership of a point owned on layer 2.  The default Urbit client reviews and validates the state change before applying the change to the Urbit's own \lstinline[style=inlinecode]{/app/azimuth}.
\end{itemize}

This mechanism means that these points are still cryptographically controlled for purposes of anyone running the default Urbit OS or relevant code (namely, \lstinline[style=inlinecode]{/app/eth-watcher}, \lstinline[style=inlinecode]{/lib/naive.hoon}, \lstinline[style=inlinecode]{/app/azimuth-tracker} and \lstinline[style=inlinecode]{/app/azimuth}), and it is easy to imagine how this same mechanism could be used to support other chains as hosts for portions of the \textsc{pki}:

\begin{enumerate}
  \item  Define a holding address for the \textsc{erc}-721 on mainnet ethereum for other host chains (i.e. sending the L1 NFT to \lstinline[style=inlinecode]{0x2222...} indicates sending the ownership to bitcoin).
  \item  Implement a watcher for the subject chain that looks for posted signature data.
  \item  Update and/or add to the relevant Urbit OS code for client side validation of data posted to the new chain.
\end{enumerate}

\noindent
This approach would increase legibility to other chains, giving communities more options for what their members can achieve with their Urbit ID.

However, as the \textsc{pki} fractures, maintaining global consistency becomes increasingly complex. In the above model, the primary solution for maintaining this consistency becomes that official `global consensus' is on mainnet Ethereum, while any other alternative is subject only to the client side validation. Depending on the affordances of the subject chain, the level of effort put into building the subsystem, and any cost concerns for network fees on the specific subject chain, the result is likely that \emph{moving to any other chain is a one-way trip}. Just as a move to the current naive rollup is a one way trip.

@@

\subsection[Move the \textsc{pki} to Nockchain]{Move the PKI to Nockchain}

While Ethereum has made significant progress in interchain bridges and L2 rollups since the implementation of the Naive rollup, making two-way bridges increasingly possible, they still introduce a host of security and consensus questions. They also are likely to exacerbate the capital outflow concerns of the current L1 Ethereum implementation.

@@

What is the best way to enable this portability (another shared principle/design goal for \textsc{ssi}s/\textsc{did}s) while maintaining global consistency? The reader who is familiar with Nockchain will foresee the punchline:  zero-knowledge proofs.

Using Zorp's \textsc{eden} and its associated Nock prover \citep{Zorp2023}, any Urbit instance can prove valid computation of any `client-side' Azimuth state change and use that proof as the data to post to any given blockchain. It is likely that the `easiest' place to do that would be Nockchain. However, external legibility may incentivize these proofs to be posted elsewhere. An Azimuth contract on a subject chain could use the proof of valid Nock computation to assign ownership of an Urbit ID crypto-asset native to a given address. This would include whatever affordances native asset ownership on that chain may bestow to an Urbit operator.

In the long run, this pathway would likely lead more Urbit operators to move their Urbit ID's primary `citizenship' to Nockchain. This would occur without unnecessarily losing legibility to other networks, and retains the possibility of absorbing external capital into the Urbit network along the way.

TODO: Mayflower? GroundWire?
At the time of press, two new proposals have been fielded which may also afford a new solution to hosting the Azimuth \textsc{pki}.  Both are in a high degree of flux and it is difficult to   Mayflower
We also note that an early design consideration of the 21e8 Bitcoin Ordinals project was to be able to host an Azimuth-like \textsc{pki} system.

\section{Other Considerations for Azimuth}

While Azimuth's on-chain nature is that of a public key infrastructure, a \textsc{pki} alone does not constitute a complete identity system. In fact, it is the pairing of Urbit ID with Urbit OS that presents the best opportunity to serve as a fully-fledged decentralized identity system. Whether we look at implementing W3C's standardized \textsc{did}s, or adhering to the principles of \textsc{ssi}s, not everything about Azimuth is, or should be, on-chain.

@@

Globally available and transparent chain state greatly enhances legibility. Openly available sponsorship hierarchies inform unconnected peers about how to reach you, and global claims can be used for various purposes. However, not every application requires global consensus, and Turing-complete smart contracts can lead to the risk of open-ended cost overruns.  Furthermore, many networks may wish to become or remain illegible to the outside world. Private access control lists, gossip networks, and packet routing mechanisms enable use cases that would not benefit from the cost or transparency of on-chain or globally consistent legibility.\footnote{It is worth noting that illegible subnets may contend with the \textsc{ssi} principle of consent, in that they enable opaque data-sharing networks and other similar `darknet' use cases.} Urbit's off-chain and encrypted peer-to-peer messaging affordances open up an intriguing landscape of `illegible subnets' while still benefiting from associations with specific pseudonymous Urbit identities.

The value of illegible subnets also builds on the insight that not everything can, or should, be distilled down to its lowest common denominator, tokenized, and put on the blockchain. The more complex the interaction, the more costly the distillation process, and \emph{the more likely there is no consensus to be found around its `valuation'}. Instead of demanding timestamped global consensus in all our peer interactions, we must support emergent, implied-consensus networks. These networks are governed not by immutable chain state but by willing association and participation by sovereign peers. Systems where the network forms and dissolves on the fly, based on the ongoing discovery and refinement of the values of its participants.

@@

The principles of \textsc{ssi} are as follows (``Ten Principles of Self-Sovereign Identity," 26 April 2016), all quoted:
